A Reduced Order Model for Permeability Fields Using Singular Value Decomposition, SVD, by Interpolating in the Principal Component Spaces \red{(Maybe you should consider a shorter and punchier title ...)}

### 2018-03-24 Papers/AppliedMaths2018/Stochastic_Upscaling_UK_AMM_2018.tex
Updates

\title{A Reduced Order Model for Permeability Fields Using Singular Value Decomposition, SVD, by Interpolating in the Principal Component Spaces}
\author[UoA]{B. Lashore\corref{cor1}}\ead{lashorebabatunde@yahoo.com} \author[UoA]{J.L.M.A. Gomes} \author[UoA]{K. Christou}
\cortext[cor1]{Corresponding author.}
\address[UoA]{Mechanics of Fluids, Soils \& Structures Group, School of Engineering, University of Aberdeen, UK}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\section{Introduction}\label{section:intro}
Naturally occuring rock formations are inherently heterogeneous, which means that the rock properties such as permeability, pore spaces ({\it i.e.}, porosity) and pore throat vary from point to point, and at different length scales. The heterogeneity of rock formations (studied under the general title of porous media) is relevant in calculations required for understanding transportation and storage of fluids in porous media. This is because the heterogeneous properties are important parameters in the mathematical models used for the aforementioned calculations. Therefore, the representation of these heterogeneous properties is of great importance to the porous media and applied mathematics community as well as the oil and gas industry, the Carbon Capture and Storage (CCS) industry, and waste management industry for whom this has direct application in reservoir management, CO$_2$ transportation and storage in subsurface rocks and remediation of contaminated soil respectively.

Furthermore, due to limitations in computing resources, for simulations with very fine grids it is neither efficient nor possible to obtain exact values of geological properties at all points in the domain (\ie scale problems) \cite{Renard_1997} \cite{miller_1998} \cite{chen_2006}. This is particularly true for large geological domains which span several kilometres. Additionally, it is not possible to obtain reliable and accurate spatial information about geological properties throughout the domain ({\it i.e.}, uncertainty problems). It is possible to overcome these challenges by upscaling. Upscaling techniques solve both problems by replacing discrete geological and fluid property values of detailed high resolution domains with coarse descriptions (low resolution) of these properties \cite{Vereecken_2007}. And upscaling techniques which best preserve the statistical properties and flow dynamics behaviour of the high resolution domain gives the best representation of the heterogeneous property.

There are several upscaling techniques which fall under different categories \cite{Renard_1997, Szymkiewicz_2013, Hasting_2001}, however over the last decade, techniques classed as stochastic have received great attention from academic and industrial porous media communities worldwide \cite{Verwoerd_2009, Ravalec-Dupin_2010, Guilleminot_2012}. One of the reasons for this is because stochastic techniques are robust enough to handle multiphase flow in porous media, and they address scale and uncertainity problems which were previously mentioned.

This work couples stochastic and deterministic upscaling representations of the permeability field to a novel high-order accurate control volume finite element method (CVFEM) (see \cite{Gomes_2017} for further details of the model formulation) so as to investigate statistical properties and multiphase flow behaviour in highly heterogeneous porous media.

Four upscaling techniques are investigated in this research. The arithmetic and harmonic mean are used to obtain the first two upscaled representation and these two are classed as deterministic techniques. The third is a randomly generated permeability field with a Guassian distribution, prescribed by a probability density function (PDF) which was obtained from a base case ({\it i.e.}, high-resolution mesh with known permeability distribution). The four upscaling technique is the crux of this research, it introduces singular value decomposition (SVD) as a method of upscaling using interpolation within the concept of principal component space to reduce the order of the permeability field.

In section 2, the basic mathematical explanation of SVD and its associated linear algebra properties are presented. The the concept of spaces is explained from the engineer's perspective and also from the mathematician's perspective. The objective of this explanation is to assist both parties in visualizing the relevant spaces and performing the necessary transformation within those spaces. Finally, the section describes model order space reduction for the permeability field within the principal component space. Section 3, starts by describing the high resolution base case on which the four upscaling techniques are modelled. The the simulation run which is similar for all the cases/models is described. A brief summary of the pre-processing step for each of the models is also presented. Section 4 provides the results and some discussion on the results of the simulations. It als gives further intepretation to the result obtained from the SVD upscaling technique. Finally, section 5 provides the conclusion.

@@

Singular Value Decomposition, SVD, exists within the linear algebra branch of mathematics as one of the most exciting means of factorizing a matrix, $\mathbf{A}$, into three parts $\mathbf{U}$, $\Sigma$ and $\mathbf{V}$. Where $\mathbf{U}$ and $\mathbf{V}$ are orthonormal, and $\Sigma$ is a diagonal matrix. If $\mathbf{A}$, is a square matrix then $\mathbf{U}$, $\Sigma$ and $\mathbf{V}$ are square matrix and $\mathbf{U}^{\intercal}\mathbf{U} = \mathbf{V}^{\intercal}\mathbf{V} = \mathbf{I}$. But if $\mathbf{A}$ is an $m * n$ matrix, $\mathbf{U}$ and $\mathbf{V}$ still remain as square matrices, albeit with different matricial sizes. $\mathbf{U}$ is an $m * m$ while $\mathbf{V}$ is an $n * n$. Another interesting property of $\mathbf{U}$ and $\mathbf{V}$ since they are square (irrespective of whether A is square or rectangular) is that $\mathbf{U}^{-1} = \mathbf{U}^{\intercal}$ and $\mathbf{V}^{-1} = \mathbf{V}^{\intercal}$. When $\mathbf{A}$ is an $m * n$ matrix, $\Sigma$ is a diagonal $m x n$ matrix. The diagonal entries of $\Sigma$ (in both the case of square and rectangular matrices) are referred to as the singular values of $\mathbf{A}$, and are usually denoted by sigma ($\sigma_{1}, \sigma_{2},..., \sigma_{r}$). For a rectangular matrix, the singular values fill the first $r$ places on the main diagonal of $\Sigma$, where $r$ is the rank of A. Additionally, the $i$th singular value, $\sigma_{i}$ is ordered such that $\sigma_{1} \geq \sigma_{2} \geq ...\geq \sigma_{r}$. A huge benefit of the SVD is the orthogonality of the columns in $\mathbf{U}$ and $\mathbf{V}$. The columns of $\mathbf{U}$ are known as the left singular vectors while the columns in $\mathbf{V}$ are known as the right singular vectors. A simple consideration of the benefit of this, is the ease with which it allows the inverse of square matrix $\mathbf{A}$ to be calculated. This example is presented thus:

>>>>

\title{A Reduced Order Model for Permeability Fields Using Singular Value Decomposition, SVD, by Interpolating in the Principal Component Spaces \red{(Maybe you should consider a shorter and punchier title ...)}}
\author[UoA]{B. Lashore\corref{cor1}}\ead{lashorebabatunde@yahoo.com} \author[UoA]{J.L.M.A. Gomes} \author[UoA]{K. Christou}
\cortext[cor1]{Corresponding author.}
\address[UoA]{Mechanics of Fluids, Soils \& Structures Research Group \\ School of Engineering University of Aberdeen, UK}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\section{Introduction}\label{section:intro}
Naturally occuring rock formations are inherently heterogeneous, which means that rock properties such as permeability, pore spaces ({\it i.e.}, porosity) and pore throat vary spatially at different length scales. The heterogeneity of rock formations (\ie porous media) is relevant in calculations of transport and storage of fluids in porous media. Therefore, the representation of these heterogeneous properties is of great importance to oil and gas (\ie reservoir management), carbon capture and storage (CCS, \ie CO$_2$ transportation and storage in subsurface rocks) and waste management (\ie remediation of contaminated soil).

\JGnote{You need to ensure that the paragraphs are somehow linked ... \eg how is the first sentence of the following paragraph linked with the previous paragraph? }

Furthermore, due to limitations in computing resources, for simulations with very fine grids it is neither efficient nor possible to obtain exact values of geological properties at all spatial coordinates of the domain (\ie scale problems) \cite{chen_2006,miller_1998,Renard_1997}. This is particularly true for large geological domains which span several kilometres. Additionally, it is not possible to obtain reliable and accurate spatial information about geological properties throughout the domain ({\it i.e.}, uncertainty problems) \red{(Isn't this sentence a repetition of the first one?)}. It is possible to overcome these challenges by upscaling. Upscaling techniques solve both problems by replacing discrete geological and fluid properties of detailed high resolution domains with coarse descriptions (low resolution) of these properties \cite{Vereecken_2007}.

Upscaling techniques which best preserve the statistical properties and flow dynamics behaviour of the high resolution domain give the best representation of the heterogeneous property. There are several upscaling techniques which fall under different categories \cite{Hasting_2001, Renard_1997, Szymkiewicz_2013}, however over the last decade, techniques classed as stochastic have received great attention from academic and industrial porous media communities worldwide \cite{Guilleminot_2012, Ravalec-Dupin_2010, Verwoerd_2009}. One of the reasons for this is because stochastic techniques are robust enough to handle multiphase flow in porous media, and they address scale and uncertainity problems which were previously mentioned.

\JGnote{The technique we are using/advocating here is not stochastic, though it would be good to develop it further, \ie brief lit review. Maybe you should add a few sentences/paragraphs on: ROM and the use of SVD for upscalling ... }
\medskip

This work couples stochastic and deterministic upscaling representations of the permeability field to a novel high-order accurate control volume finite element method (CVFEM) (see \cite{Gomes_2017} for further details of the model formulation) so as to investigate statistical properties and multiphase flow behaviour in highly heterogeneous porous media. \red{(Try to split this really long sentence ...)}

Four upscaling techniques are investigated in this work. Arithmetic and harmonic means are used to obtain the first two upscaled representation and these two are classed as deterministic techniques. The third is a randomly generated permeability field with a Gaussian distribution, prescribed by a probability density function (PDF) which was obtained from a domain discretised with high-resolution mesh with known permeability distribution (\ie  base case). The fourth upscaling technique is the crux of this research, it introduces singular value decomposition (SVD) as a method of upscaling using interpolation within the concept of principal component space to reduce the order of the permeability field \red{(Sentence is quite confusing and does not say much ...)}.

In Section 2, the basic mathematical explanation of SVD and its associated properties are presented. Then, the concept of spaces is explained from the engineer's perspective and also from the mathematician's perspective. The objective of this explanation is to assist both parties in visualizing the relevant spaces and performing the necessary transformation within those spaces. Finally, the section describes model order space reduction for the permeability field within the principal component space. Section 3, starts by describing the high resolution base case on which the four upscaling techniques are modelled. The the simulation run which is similar for all the cases/models is described. A brief summary of the pre-processing step for each of the models is also presented. Section 4 provides the results and some discussion on the results of the simulations. It als gives further intepretation to the result obtained from the SVD upscaling technique. Finally, section 5 provides the conclusion.

@@

Singular Value Decomposition, SVD, \JGnote{ioioioiok} exists within the linear algebra branch of mathematics as one of the most exciting means of factorizing a matrix, $\mathbf{A}$, into three parts $\mathbf{U}$, $\Sigma$ and $\mathbf{V}$. Where $\mathbf{U}$ and $\mathbf{V}$ are orthonormal, and $\Sigma$ is a diagonal matrix. If $\mathbf{A}$, is a square matrix then $\mathbf{U}$, $\Sigma$ and $\mathbf{V}$ are square matrix and $\mathbf{U}^{\intercal}\mathbf{U} = \mathbf{V}^{\intercal}\mathbf{V} = \mathbf{I}$. But if $\mathbf{A}$ is an $m * n$ matrix, $\mathbf{U}$ and $\mathbf{V}$ still remain as square matrices, albeit with different matricial sizes. $\mathbf{U}$ is an $m * m$ while $\mathbf{V}$ is an $n * n$. Another interesting property of $\mathbf{U}$ and $\mathbf{V}$ since they are square (irrespective of whether A is square or rectangular) is that $\mathbf{U}^{-1} = \mathbf{U}^{\intercal}$ and $\mathbf{V}^{-1} = \mathbf{V}^{\intercal}$. When $\mathbf{A}$ is an $m * n$ matrix, $\Sigma$ is a diagonal $m x n$ matrix. The diagonal entries of $\Sigma$ (in both the case of square and rectangular matrices) are referred to as the singular values of $\mathbf{A}$, and are usually denoted by sigma ($\sigma_{1}, \sigma_{2},..., \sigma_{r}$). For a rectangular matrix, the singular values fill the first $r$ places on the main diagonal of $\Sigma$, where $r$ is the rank of A. Additionally, the $i$th singular value, $\sigma_{i}$ is ordered such that $\sigma_{1} \geq \sigma_{2} \geq ...\geq \sigma_{r}$. A huge benefit of the SVD is the orthogonality of the columns in $\mathbf{U}$ and $\mathbf{V}$. The columns of $\mathbf{U}$ are known as the left singular vectors while the columns in $\mathbf{V}$ are known as the right singular vectors. A simple consideration of the benefit of this, is the ease with which it allows the inverse of square matrix $\mathbf{A}$ to be calculated. This example is presented thus:
